worldmap

### 2021-09-22 analysis/exporatory_scripts/BAT_explore.Rmd

>>>>

Let's also do the functions from BAT package
```{r}
#function for getting PCA axes with at least 95% coverage
axes95 <- function(mat){
  axes <- pco(mat)
  eigen <- axes$eig

  threshold <- sum(eigen[eigen > 0]) * 0.95

  sums <- cumsum(eigen)
  num_axes <- Position(function(x) x > threshold, sums)

  return(num_axes)
}

get_BAT_fd <- function(data, filter_plot){
  require(BAT)
  require(StatMatch)
  require(labdsv)

  data <- data %>% filter(plot_id == filter_plot)

  print(max(data$record_id))

  species_mat <- data %>%
    select(year, id) %>% 
    distinct() %>%
    arrange(year, id) %>%
    mutate(present = 1) %>%
    spread(id, value = present, fill = 0) %>%
    column_to_rownames("year")

  trait_mat <- data %>%
    select(-c(taxa, record_id, day, month, year, sample_desc, id_species, 
              latitude, longitude, sum_abundance, sum_biomass, biotimeName, 
              traitName, scientificName)
    ) %>%
    distinct() %>%
    select_if(~ length(unique(.)) > 1) %>% #remove columns that have the same value for all columns
    mutate_at(vars(-id, -ends_with("5cat")), list(~scales::rescale(.,to=c(0, 1)))) %>% #rescale variables
    arrange(id) %>%
    column_to_rownames("id")

  #get PCA trait matrix
  trait_dist <- gower.dist(trait_mat)
  trait_pco <- pco(trait_dist, k = axes95(trait_dist))

  traits <- trait_pco$points
  rownames(traits) <- rownames(trait_mat)

  #getFD metrics
  kernelFD <- kernel.alpha(comm=as.matrix(species_mat),trait=traits,abund=FALSE,method="gaussian",return.hv=TRUE)
  hv <- kernelFD[[2]]

  disp <- kernel.dispersion(comm=hv,frac=0.1,func='dissimilarity') 
  div <- kernel.dispersion(comm=hv,frac=0.6, func='divergence')
  even <- kernel.evenness(comm=hv)

  results <- data.frame(richness = as.numeric(kernelFD[[1]]), disp, div, even) %>% 
    mutate(plot_id = filter_plot)

  return(results)
}

get_BAT_fd(biotime_traits, 376300153)
```

### 2021-09-22 analysis/exporatory_scripts/problem_model.Rmd

metrics <- pins::pin_get("rarefied-metrics", board = "github") %>%
  # pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
  filter(!is.na(value)) %>%
  mutate(logvalue = log(value + 1),
         year_scaled = scale(year)) # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values

cont_inf <- metrics %>% filter(is.infinite(value), !metric %in% c("qual.FRic",  "SES_qual.FRic")) %>% pull(rarefyID)

@@

model_data <- metrics %>% 
  filter(!rarefyID %in% cont_inf, !rarefyID %in% missing_null) %>%
  mutate(study_id = substr(rarefyID, 1, 3))
```

@@

Visualize distributions of response
```{r}
model_data %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S_int")) %>%
  ggplot() +
  geom_histogram(aes(x = logvalue),
                 bins = 60) +
  facet_wrap(~metric, scales = "free") +
  labs(title = "Distribution of Log values")

model_data %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S_int")) %>%
  ggplot() +
  geom_histogram(aes(x = value),
                 bins = 60) +
  facet_wrap(~metric, scales = "free") +
  labs(title = "Distribution of raw values")
```

Fit a single model to play around with 
```{r}
test_data <- model_data %>% 
  filter(metric == "FEve") %>%
  mutate(study_id = as.factor(study_id))

#lognormal
ln_fit <- lmer(logvalue ~ year_center + (year_center|study_id/rarefyID), data = test_data,
               control = lmerControl(optCtrl = list(maxit = 1e9,
                                                    maxfun = 1e9,
                                                    xtol_abs = 1e-11,
                                                    ftol_abs = 1e-11,
                                                    maxeval = 1e9)))

@@

```{r}
options(mc.cores = parallel::detectCores())
stan_fit <- stan_lmer(value_scale ~ year_center + (year_center|study_id/rarefyID), data = test_data, iter = 500)

launch_shinystan(stan_fit,ppd = FALSE)
```

>>>>

#get metadata
meta <- pins::pin_get("meta", board = "github") %>%
  select(study_id, realm, climate, habitat, protected_area, biome_map)

meta <- pins::pin_get("bt-traitfiltered", board = "github") %>%
  select(study_id, rarefyid, taxa) %>% distinct() %>%
  left_join(meta)

#get metric data
metrics <- pins::pin_get("rarefied-metrics", board = "github") %>%
  # pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
  filter(!is.na(value)) %>%
  mutate(logvalue = log(value + 1)) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
  left_join(meta, c("rarefyID" = "rarefyid"))

#find rarefyID's that have inf value for some metrics and exclude
cont_inf <- metrics %>% filter(is.infinite(value), !metric %in% c("qual.FRic",  "SES_qual.FRic")) %>% pull(rarefyID)

@@

model_data <- metrics %>% 
  filter(!rarefyID %in% cont_inf, !rarefyID %in% missing_null) %>%
  group_by(metric) %>%
  mutate(study_id =  str_extract(rarefyID, "[^_]+")) %>%
  mutate(year_scaled = scale(year), 
         scale_center = attributes(year_scaled)$`scaled:center`, 
         scale = attributes(year_scaled)$`scaled:scale`) %>%
  ungroup()
```

@@

Visualize distributions of response
```{r}
model_data %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "Jaccard_base")) %>%
  ggplot() +
  geom_histogram(aes(x = logvalue),
                 bins = 60) +
  facet_wrap(~metric, scales = "free") +
  labs(title = "Distribution of Log values")

model_data %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "Jaccard_base")) %>%
  ggplot() +
  geom_histogram(aes(x = value),
                 bins = 60) +
  facet_wrap(~metric, scales = "free") +
  labs(title = "Distribution of raw values")
```

Fit a single model to play around with 
```{r}
test_data <- model_data %>% 
  filter(metric == "FRic") %>%
  mutate(study_id = as.factor(study_id))

#lognormal
ln_fit <- lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), data = test_data,
               control = lmerControl(optCtrl = list(maxit = 1e9,
                                                    maxfun = 1e9,
                                                    xtol_abs = 1e-11,
                                                    ftol_abs = 1e-11,
                                                    maxeval = 1e9)))

@@

Get p-values for blups 
```{r}
cV <- ranef(ln_fit, effects = "ran_vals", drop = TRUE)
ranvar <- attr(cV[[1]], "postVar")

ng <- dim(ranvar)[3]
np <- dim(ranvar)[2]
mm <- matrix(ranvar[cbind(rep(seq(np),ng),
             rep(seq(np),ng),
             rep(ng,each=np))],
       byrow=TRUE,
       nrow=ng)

sumVar <- vcov(ln_fit)[1,1]+mm[,1]

year_scaled_var <- tidy(ln_fit) %>%
  filter(term == "year_scaled") %>% 
  mutate(var = std.error*std.error) %>%
  pull(var)

study_ests <- broom.mixed::tidy(ln_fit, effects="ran_vals") %>%
  filter(group == "study_id", term == "year_scaled") %>% 
  rename(cond.std.error = std.error) %>%
  mutate(cond.var = cond.std.error*cond.std.error, 
         var = cond.var + year_scaled_var,
         std.error = sqrt(var),
         upr.ci = estimate + (2*std.error),
         lwr.ci = estimate - (2*std.error),
         sig = case_when(
           lwr.ci < 0 & upr.ci > 0 ~ FALSE,
           TRUE ~ TRUE
         ))
```

### 2021-09-22 analysis/figures/visualization.Rmd

ggsave(here::here("figures/study_map.jpeg"))
```

@@

ggsave2(here::here("figures/study_map_hist.jpeg"))
```

@@

group_colors <- c("Global" = my_pal[1], "Polar/Temperate" = my_pal[5], "Temperate/Tropical" = my_pal[3], "Tropical" = my_pal[4], "Temperate" = my_pal[2], "Overall Mean" = "#000000")

@@

ggsave2(here::here("figures", "3met_long.jpeg"))
```

>>>>

ggsave(here::here("figures/study_map.jpeg"), bg = "white")
```

@@

ggsave2(here::here("figures/study_map_hist.jpeg"), bg = "white")
```

@@

group_colors <- c("Global" = my_pal[1], "Polar/Temperate" = my_pal[6], "Temperate/Tropical" = my_pal[3], "Tropical" = my_pal[4], "Temperate" = my_pal[2], "Overall Mean" = "#000000")

@@

ggsave2(here::here("figures", "3met_long.jpeg"), bg = "white")
```
